### 2025-02-07 mixed-models.qmd
20250207 - clarifications

In this chapter, we use mixed models to demonstrate how to model trajectories (growth curves) of players' fantasy points as a function of age.
One of the challenges when modeling players' fantasy points as a function of age is that fantasy points are a function of both ability and opportunity.
With age, the player's ability and opportunity may both decline, but it is difficult to disentangle the extent to which a player's decline is related to declines in ability versus opportunity.
As players get older, they may be supplanted by younger, more talented players.
With age, despite still having latent (unobserved) ability, players will get fewer and, eventually, no opportunity.
And we do not know the counterfactual—how many fantasy points they would have scored had they been given the full opportunity each season.
Thus, players may go from scoring 100+ points in a season to all of a sudden scoring way fewer points, with little in the way of intermediate steps.
Thus, this should inform how you interpret the resulting trajectories.
For instance, the steep declines can make it look like the model-implied fantasy points go well below zero at later ages, even though it is obviously not likely for players to obtain such a magnitude of negative fantasy points.
The negative model-implied values of fantasy points at later ages are likely an artifact of the decreasing opportunity that players tend to get as they age.

@@

A plot of Quarterbacks' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeQB.
The model-implied peak of Quarterbacks' fantasy points is at age `{r} qbPeakAge`.
The model-predicted value of zero fantasy points for Quarterbacks is at `{r} qbZeroAge`.

@@

A plot of Fullbacks' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeFB.
The model-implied peak of Fullbacks' fantasy points is at age `{r} fbPeakAge`.
The model-predicted value of zero fantasy points for Fullbacks is at `{r} fbZeroAge`.

@@

A plot of Running Backs' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeRB.
The model-implied peak of Running Backs' fantasy points is at age `{r} rbPeakAge`.
The model-predicted value of zero fantasy points for Running Backs is at `{r} rbZeroAge`.

@@

A plot of Wide Receivers' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeWR.
The model-implied peaks of Wide Receivers' fantasy points are at ages `{r} wrPeakAge` and `{r} wrPeakAge2`.
The model-predicted value of zero fantasy points for Wide Receivers is at `{r} wrZeroAge`.

@@

A plot of Tight Ends' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeTE.
The model-implied peak of Tight Ends' fantasy points is at age `{r} tePeakAge`.
The model-predicted value of zero fantasy points for Tight Ends is at `{r} teZeroAge`.

>>>>

In this chapter, we use mixed models to demonstrate how to model trajectories (growth curves) of players' fantasy points as a function of age.
One of the challenges when modeling players' fantasy points as a function of age is that fantasy points are a function of both ability and opportunity.
With age, the player's ability and opportunity may both decline, but it is difficult to disentangle the extent to which a player's decline is related to declines in ability versus opportunity.
As players get older, they may be supplanted by younger, more talented players.
With age, despite still having latent (unobserved) ability, players will get fewer and, eventually, no opportunity.
And we do not know the counterfactual—how many fantasy points they would have scored had they been given the full opportunity each season.
Thus, players may go from scoring 100+ points in a season to all of a sudden scoring way fewer points, with little in the way of intermediate steps.
Thus, this should inform how you interpret the resulting trajectories.
For instance, the steep declines can make it look like the model-implied fantasy points go well below zero at later ages, even though it is obviously not likely for players to obtain such a magnitude of negative fantasy points.
The negative model-implied values of fantasy points at later ages are likely an artifact of the decreasing opportunity that players tend to get as they age.
The crossover point, where the positive values becomes negative might indicate, for instance, that players at that age tend to have retired and thus have no opportunity.
Regardless, the form/shape/steepness of the decline is still potentially meaningful even if the precise negative number at a given age is not.

@@

A plot of Quarterbacks' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeQB.
The model-implied peak of Quarterbacks' fantasy points is at age `{r} qbPeakAge`.
The model-predicted value of zero fantasy points for Quarterbacks is at age `{r} qbZeroAge`.

@@

A plot of Fullbacks' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeFB.
The model-implied peak of Fullbacks' fantasy points is at age `{r} fbPeakAge`.
The model-predicted value of zero fantasy points for Fullbacks is at age `{r} fbZeroAge`.

@@

A plot of Running Backs' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeRB.
The model-implied peak of Running Backs' fantasy points is at age `{r} rbPeakAge`.
The model-predicted value of zero fantasy points for Running Backs is at age `{r} rbZeroAge`.

@@

A plot of Wide Receivers' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeWR.
The model-implied peaks of Wide Receivers' fantasy points are at ages `{r} wrPeakAge` and `{r} wrPeakAge2`.
The model-predicted value of zero fantasy points for Wide Receivers is at age `{r} wrZeroAge`.

@@

A plot of Tight Ends' model-implied fantasy points by age is in @fig-individualFantasyPointsByAgeTE.
The model-implied peak of Tight Ends' fantasy points is at age `{r} tePeakAge`.
The model-predicted value of zero fantasy points for Tight Ends is at age `{r} teZeroAge`.
